---
title: "Getting Started with bboutools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bboutools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(bboutools)
```

`bboutools` is an R package to estimate Boreal Caribou recruitment, survival and population growth. 
Functions are provided to fit Bayesian or Maximum Likelihood (ML) models and generate and plot predictions.  

Under the hood, the [Nimble](https://r-nimble.org) R package is used to fit heirarchical Bayesian and Maximum Likelihood models. 
Model templates in Nimble use BUGS-like syntax.  

Several anonymized data sets are provided in a separate R package, `bboudata`. 

In order to use `bboutools` you need to install R (see below) or use the Shiny [app](https://github.com/poissonconsulting/bboushiny).

## Philosophy

`bboutools` is intended to be used in conjunction with [tidyverse](https://www.tidyverse.org) packages such as `readr`, `dplyr` to manipulate data and `ggplot2` [@ggplot2] to plot data.
As such, it endeavors to fulfill the tidyverse [manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html).

## Installation

In order to install R [@r] the appropriate binary for the users operating system should be downloaded from [CRAN](https://cran.r-project.org) and then installed.

Once R is installed, the `bboutools` package can be installed from GitHub by executing the following code at the R console
```{r, eval=FALSE}
install.packages("pak")
pak::pak("poissonconsulting/bboutools")
```

The `bboutools` package can then be loaded into the current session using
```{r, message = FALSE}
library(bboutools)
```

## Getting Help

To get additional information on a particular function just type `?` followed by the name of the function at the R console.
For example `?bb_fit_recruitment` brings up the R documentation for the `bboutools` recruitment model fit function.

For more information on using R the reader is referred to [R for Data Science](https://r4ds.had.co.nz) [@wickham_r_2016].

If you discover a bug in `bboutools` please file an issue with a [reprex](https://reprex.tidyverse.org/articles/reprex-dos-and-donts.html) (reproducible example) at <https://github.com/poissonconsulting/bboutools/issues>.

## Providing Data

Once the `bboutools` package has been loaded, the next task is to provide some data.
An easy way to do this is to create a comma separated file (`.csv`) with spreadsheet software. Recruitment and survival data should be formatted as in the following anonymized datasets:

```{r}
# Recruitment data
bboudata::bbourecruit_a
```

```{r}
# Survival data
bboudata::bbousurv_a
```

All columns should be included and column names should not be changed. 

The `.csv` file can then be read into R using the following
```{r, eval=FALSE}
data <- read_csv(file = "path/to/file.csv")
```

## Recruitment

The annual recruitment in boreal caribou population is typically estimated from annual calf:cow ratios.

`bboutools` fits a Poisson annual recruitment model to the counts of calfs, cows, yearlings unknown adults and potentially, bulls.  

It is up to the user to ensure that the data are from surveys that were conducted at the same time of year when calf survival is expected to be similar to adult survival. 

### Fit a Bayesian model

The function `bb_fit_recruitment()` fits a Bayesian recruitment model.  

The start month of the biological year (i.e., 'caribou year') can be set with the `year_start` argument.
By default, the start month is April. 
 
The adult female proportion can either be fixed or estimated from counts of cows and bulls (i.e., `Cows ~ Binomial(adult_female_proportion, Cows + Bulls)`). 

If the user provides a value to `adult_female_proportion`, it is fixed.
The default value is 0.65, which accounts for higher mortality of males [@smith_2004].
If `adult_female_proportion = NULL`, the adult female proportion is estimated.
By default, a biologically informative prior of `Beta(65,35)` is used. 
This corresponds to an expected value of 65%.  

The yearling female proportion can be set with `yearling_female_proportion`. The default value is 0.5.  

The model can be fit with random effect of year, fixed effect of year and/or continuous effect of year (i.e., year trend). 
The `min_random_year` argument dictates the minimum number of years in the dataset required to fit a random year effect; otherwise a fixed year effect is fit. 
It is not recommended to fit a random year effect with fewer than 5 years. 
A continuous fixed effect of year can be fit with `year_trend = TRUE`. 

The user can set `quiet = FALSE` argument to see messages and sampling progress. 

```{r}
recruitment <- bb_fit_recruitment(bboudata::bbourecruit_a, year_start = 4, year_trend = TRUE, quiet = TRUE)
```

### Convergence
Model convergence can be checked with the `glance()` function. 
```{r}
glance(recruitment)
```

Model convergence provides an indication of whether the parameter estimates are reliable. 

Convergence is successful if `ess` > 33% of the number of iterations and `rhat` < 1.1. 
`ess` (Effective Sample Size) represents the length of a chain (i.e., number of iterations) if each sample was independent of the one before it. 
`rhat` evaluates whether the chains agree on the same values. 
As the total variance of all the chains shrinks to the average variance within chains, r-hat approaches 1.  

To improve convergence (i.e., by increasing `ess` and/or decreasing `rhat`), the user can increase the thinning rate with the `nthin` argument.

### Summary
Various generic functions in `bboutools` can be used to summarize or interrogate the output of model fitting functions.

- `coef()` and `tidy()` provide a tidy table of the coefficient estimates. 
- `estimates()` provides a list of the coefficient estimates. 
- `augment()` provides the data used. 
- `model_code()` provides the model code in BUGS-like syntax. 
- `plot()` provides traceplots for individual parameters. 

The user can exclude individual random effect estimates from coefficient output.
```{r}
tidy(recruitment, include_random_effects = FALSE)
```

### Priors
In general, weakly informative priors are used by default [@gelman_prior_2017, @mcelreath_statistical_2016]. 
The user can change the priors by providing a named vector to `priors`. 
For example, an uninformative prior for adult female proportion (e.g., Beta distribution with alpha and beta parameter of 1), can be supplied to `priors` as follows. 
```{r, eval=FALSE}
recruitment <- bb_fit_recruitment(bboudata::bbourecruit_a, priors = c(adult_female_proportion_alpha = 1, adult_female_proportion_beta = 1))
```

## Survival
The annual survival in boreal caribou population is typically estimated from the monthly fates of collared adult females. 
`bboutools` fits a Binomial monthly survival model to the number of collared females and mortalities.
The user can choose whether to include individuals with uncertain fates with the certain mortalities.  

The function `bb_fit_survival()` fits a Bayesian survival model. 

The survival model is always fit with a random intercept for each month. 
Otherwise, the `year_start`, `year_trend`, and `min_random_year` arguments have the same behaviour as `bb_fit_recruitment()` above. 

If `include_uncertain_mortalities = TRUE`, the total mortalities is the sum of the certain mortalities and uncertain mortalities ('MortalitiesCertain' and 'MortalitiesUncertain' columns); otherwise, only certain mortalities are used to fit the model. 

```{r}
survival <- bb_fit_survival(bboudata::bbousurv_a, year_start = 4, quiet = TRUE)
```

```{r}
tidy(survival, include_random_effects = FALSE)
```

## Predictions
A user can generate and plot predictions of recruitment, survival and population growth. 

#### Recruitment by year
```{r}
predict_recruitment <- bb_predict_recruitment(recruitment, year = TRUE)
bb_plot_year_recruitment(predict_recruitment)
```

#### Recruitment for a 'typical' year
```{r}
predict_recruitment_1 <- bb_predict_recruitment(recruitment, year = FALSE)
predict_recruitment_1
```

#### Recruitment trend
```{r}
predict_recruitment_trend <- bb_predict_recruitment_trend(recruitment)
bb_plot_year_trend_recruitment(predict_recruitment_trend)
```

#### Survival by year for a 'typical' month
```{r}
predict_survival <- bb_predict_survival(survival, year = TRUE, month = FALSE)
bb_plot_year_survival(predict_survival)
```

#### Survival by month for a 'typical' year

The estimates show annual survival, i.e., if that month lasted the duration of the year. 

```{r}
predict_survival_month <- bb_predict_survival(survival, year = FALSE, month = TRUE)
bb_plot_month_survival(predict_survival_month)
```

## Population Growth

The rate of population growth ($\lambda$) is

$$\lambda = \frac{N_{t+1}}{N_t}$$

where $N_t$ is the population abundance in year $t$.

Following @hatter_moose_1991, it can be shown that

$$\lambda = \frac{S}{1-R}$$

As ungulate populations are generally polygynous survival and recruitment are estimated with respect to the number of adult (mature) females.

Following @decesare_estimating_2012, before calculating $\lambda$ the annual recruitment is first divided by two to give the expected number of female calves per adult female (under the assumption of a 1:1 sex ratio).

$$R_F = R/2$$

Next the annual recruitment is adjusted to give the proportional change in the population.

$$R_\Delta =  \frac{R_F}{1 + R_F}$$

Finally $\lambda$ is calculated using

$$\lambda = \frac{S}{1-R_\Delta}$$
A user can predict population growth with `bb_predict_growth()`, feeding in the previously fit survival and recruitment models. 
```{r}
predict_lambda <- bb_predict_growth(survival = survival, recruitment = recruitment)
bb_plot_year_growth(predict_lambda) +
  ggplot2::scale_y_continuous(labels = scales::percent)
```

Population change (%) is calculated with uncertainty as the cumulative product of population growth. 

```{r}
predict_change <- bb_predict_population_change(survival = survival, recruitment = recruitment)
bb_plot_year_population_change(predict_change)
```

### Maximum Likelihood
Maximum Likelihood (ML) models can be fit using the `bb_fit_recruitment_ml()` and `bb_fit_survival_ml()` functions. 
These functions take a few seconds to execute because Nimble must compile the model into C++ code.
See the [Nimble documentation](https://r-nimble.org/html_manual/cha-AD.html#how-to-use-laplace-approximation) for more information and comparison to TMB.
Similar to Bayesian model fits, generic functions such as `tidy()`, `glance()` and `augment()` work on ML fit objects (class 'bboufit_ml').

```{r}
recruitment_ml <- bb_fit_recruitment_ml(bboudata::bbourecruit_a, year_start = 4, year_trend = TRUE, quiet = TRUE)
```

```{r}
glance(recruitment_ml)
```

```{r}
survival_ml <- bb_fit_survival_ml(bboudata::bbousurv_a, year_start = 4, quiet = TRUE)
```

The ML estimates are comparable to estimates derived from the equivalent Bayesian models above. In general, ML models can be interpreted as Bayesian models with uninformative (e.g., uniform) priors [@mcelreath_statistical_2016].

```{r}
tidy(recruitment_ml, include_random_effects = FALSE)
```

```{r}
tidy(survival_ml, include_random_effects = FALSE)
```

There is no functionality in `bboutools` to generate predictions (i.e., derived parameters) from ML models. 
This is a more straightforward task with Bayesian models. 

Note that ML models can struggle to converge with a fixed year effect if there are many years in the dataset. 
Insufficient information to estimate each year independently results in a flat likelihood surface and inability of the model to reliably identify the ML estimates.
This can be resolved by estimating year as a random effect, continuous effect (`year_trend`) or excluding year altogether (i.e., with `exclude_year = TRUE`).

Another possible source of convergence issues is initial values. 
By default, `bboutools` sets initial values based on default priors used for parameters in the Bayesian models. 
The user can replace initial values for parameters using `inits`.

```{r}
inits_ml <- bb_fit_recruitment_ml(bboudata::bbourecruit_a, inits = c(b0 = 1, sAnnual = 0.3))
```

## Going beyond `bboutools`
The output of `bb_fit_recruitment()` and `bb_fit_survival()` is a named list with class `bboufit`. It contains the MCMC samples (class `mcmcr`), data, model code, and Nimble model object. 

```{r}
names(recruitment)
```

A user can do more than the existing `bboutools` functionality provides by working directly with these objects. 
Packages such as `mcmcr`, `mcmcderive` and `coda` can be used to work with MCMC samples. 

The [Nimble](https://r-nimble.org) R package can be used to work directly with the Nimble model object.

## References