---
title: "Prior Selection and Influence"
bibliography: bibliography.bib
---

<!-- # Copyright 2022-2023 Integrated Ecological Research and Poisson Consulting Ltd. -->
<!-- # Copyright 2024 Province of Alberta -->
<!-- # -->
<!-- # Licensed under the Apache License, Version 2.0 (the "License"); -->
<!-- # you may not use this file except in compliance with the License. -->
<!-- # You may obtain a copy of the License at -->
<!-- # -->
<!-- # http://www.apache.org/licenses/LICENSE-2.0 -->
<!-- # -->
<!-- # Unless required by applicable law or agreed to in writing, software -->
<!-- # distributed under the License is distributed on an "AS IS" BASIS, -->
<!-- # WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied. -->
<!-- # See the License for the specific language governing permissions and -->
<!-- # limitations under the License. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

In Bayesian analysis, a prior distribution encodes what is known or assumed about a parameter before observing the data.
The prior is combined with the likelihood of the data to produce the posterior distribution, which represents the updated knowledge about the parameter after observing the data.
When the data are abundant, the posterior is dominated by the likelihood and the choice of prior has little practical effect.
When the data are sparse, the prior has more influence on the posterior, which can be beneficial if the prior reflects genuine biological knowledge, or problematic if the prior is poorly chosen.

Priors range from uninformative (assigning equal probability to all values) to strongly informative (concentrating probability near a specific value).
`bboutools` uses weakly informative priors by default, which rule out biologically implausible values but remain diffuse enough to let the data drive the estimates [@gelman_prior_2017; @mcelreath_statistical_2016].

If the user is interested in fitting models without priors, see `bb_fit_recruitment_ml()` and `bb_fit_survival_ml()`, which have identical models but use a frequentist approach (Maximum Likelihood) to parameter estimation.
Models fit with Maximum Likelihood are equivalent to Bayesian models with completely uninformative priors [@mcelreath_statistical_2016].

### Survival

Given the full model, the expected survival probability for the $i^{th}$ year and $j^{th}$ month is
$$\text{logit}(Survival[i,j]) = b0 + bAnnual[i] + bMonth[j] + bYear \cdot Year[i]$$
Where $bAnnual$ can be a fixed or random effect of categorical year on the intercept on the log-odds scale and $Year$ is the scaled continuous year.  

This model has the following default priors in `bboutools`

- Intercept (log-odds scale)
$$b0 \sim Normal(3, sd = 10)$$
- Year fixed effect
$$bAnnual[i] \sim Normal(0, sd = 10)$$
- Year random effect
$$sAnnual \sim Exponential(1)$$
$$bAnnual \sim Normal(0, sd = sAnnual)$$

- Month random effect
$$sMonth \sim Exponential(1)$$
$$bMonth \sim Normal(0, sd = sMonth)$$

- Year continuous effect
$$bYear \sim Normal(0, sd = 2)$$

### Recruitment

Given the full model, the expected recruitment (calves per adult female) for the $i^{th}$ year is
$$logit(Recruitment[i]) = b0 + bAnnual[i] + bYear \cdot Year[i]$$
where $bAnnual$ can be a fixed or random effect of categorical year on the intercept on the log scale and $Year$ is the scaled continuous year. 

The model has the following default priors in `bboutools`

- Intercept (log-odds scale)
$$b0 \sim Normal(-1, sd = 5)$$

- Categorical year fixed effect
$$bAnnual[i] \sim Normal(0, sd = 5)$$

- Categorical year random effect
$$sAnnual \sim Exponential(1)$$
$$bAnnual \sim Normal(0, sd = sAnnual)$$

- Continuous year effect
$$bYear \sim Normal(0, sd = 2)$$ 


The recruitment model can also estimate the adult female proportion from observed Cows and Bulls
$$Cows = adult\_female\_proportion \cdot (Cows + Bulls)$$
Where the default prior for $adult\_female\_proportion$ has a mode of 65%
$$adult\_female\_proportion \sim Beta(65, 35)$$  

### Prior selection and influence

The default weakly informative priors are appropriate for most analyses.
A user may wish to adjust priors in several situations: when strong biological knowledge exists for a parameter (e.g., adult female proportion from published studies), when the dataset is small and the user wants to constrain estimates to a plausible range, or when comparing the sensitivity of results to different prior assumptions.
The examples below illustrate how tightening or relaxing priors affects parameter estimates relative to the Maximum Likelihood baseline.

```{r setup}
library(bboutools)
library(dplyr)
library(ggplot2)
```

#### Adult female proportion

To illustrate the effect, we construct a dataset where a large proportion of adults are unclassified by moving half of observed Cows and Bulls into `UnknownAdults`.
We then compare recruitment estimates from priors centered on high (80%) and low (35%) adult female proportions, along with the default and maximum likelihood estimates.

```{r}
set.seed(1)
data <- bboudata::bbourecruit_a %>%
  filter(Year >= 2005, Year <= 2010) %>%
  mutate(
    UnknownAdults = UnknownAdults + round(Cows / 2) + round(Bulls / 2),
    Cows = Cows - round(Cows / 2),
    Bulls = Bulls - round(Bulls / 2)
  )

# default prior: Beta(65, 35), mode at 65%
fit <- bb_fit_recruitment(data, adult_female_proportion = NULL, quiet = TRUE)

# high prior: Beta(80, 20), mode at ~81%
fit_high <- bb_fit_recruitment(
  data,
  adult_female_proportion = NULL,
  quiet = TRUE,
  priors = c(
    "adult_female_proportion_alpha" = 80,
    "adult_female_proportion_beta" = 20
  )
)

# low prior: Beta(35, 65), mode at ~35%
fit_low <- bb_fit_recruitment(
  data,
  adult_female_proportion = NULL,
  quiet = TRUE,
  priors = c(
    "adult_female_proportion_alpha" = 35,
    "adult_female_proportion_beta" = 65
  )
)

# maximum likelihood for comparison
fit_ml <- bb_fit_recruitment_ml(
  data,
  adult_female_proportion = NULL,
  quiet = TRUE
)
```

The choice of prior on $adult\_female\_proportion$ affects the predicted recruitment.

```{r, echo=FALSE,fig.width=6,fig.height=4}
pred <- bb_predict_recruitment(fit)
pred_high <- bb_predict_recruitment(fit_high)
pred_low <- bb_predict_recruitment(fit_low)
pred_ml <- bb_predict_recruitment(fit_ml)

df <- bind_rows(
  list(
    "Default Beta(65,35)" = pred,
    "High Beta(80,20)" = pred_high,
    "Low Beta(35,65)" = pred_low,
    "Maximum Likelihood" = pred_ml
  ),
  .id = "model"
)

ggplot(data = df) +
  geom_pointrange(
    aes(
      x = factor(CaribouYear),
      y = estimate,
      ymin = lower,
      ymax = upper,
      color = model
    ),
    position = position_dodge(width = 0.7)
  ) +
  theme(legend.position = "bottom") +
  scale_y_continuous("Recruitment") +
  xlab("Caribou Year") +
  scale_color_discrete("")
```

The influence of the $adult\_female\_proportion$ prior on recruitment predictions depends on the proportion of unclassified adults in the data.
When most adults are already classified as Cows or Bulls, the prior has little practical effect on recruitment.

The $adult\_female\_proportion$ can also simply be fixed.
See `bb_fit_recruitment()` for details.

#### Survival intercept

A user may have prior knowledge about the expected survival rate for their population from published studies or neighboring populations.
This can be encoded in the intercept prior `b0`.
The default prior for survival is `b0 ~ Normal(3, 10)` on the logit scale, which is very diffuse.

As an example, suppose published estimates suggest annual survival is approximately 85% for a population.
On the logit scale, 85% annual survival corresponds to a monthly survival of approximately $0.85^{1/12} \approx 0.9865$, or $\text{logit}(0.9865) \approx 4.3$.
A prior of `b0 ~ Normal(4.3, 0.5)` would encode this belief while still allowing the data to shift the estimate.

```{r}
set.seed(1)
data <- bboudata::bbousurv_c

# default prior: b0 ~ Normal(3, 10)
fit_default <- bb_fit_survival(data, quiet = TRUE)

# informative prior centered on ~85% annual survival
fit_informed <- bb_fit_survival(
  data,
  priors = c(b0_mu = 4.3, b0_sd = 0.5),
  quiet = TRUE
)

# maximum likelihood for comparison
fit_ml <- bb_fit_survival_ml(data, quiet = TRUE)
```

```{r, echo=FALSE,fig.width=6,fig.height=4}
pred_default <- bb_predict_survival(fit_default)
pred_informed <- bb_predict_survival(fit_informed)
pred_ml <- bb_predict_survival(fit_ml)

df <- bind_rows(
  list(
    "Default Normal(3, 10)" = pred_default,
    "Informed Normal(4.3, 0.5)" = pred_informed,
    "Maximum Likelihood" = pred_ml
  ),
  .id = "model"
)

ggplot(data = df) +
  geom_pointrange(
    aes(
      x = factor(CaribouYear),
      y = estimate,
      ymin = lower,
      ymax = upper,
      color = model
    ),
    position = position_dodge(width = 0.5)
  ) +
  theme(legend.position = "bottom") +
  scale_y_continuous("Annual Survival", labels = scales::percent) +
  xlab("Caribou Year") +
  scale_color_discrete("")
```

The informative prior pulls survival estimates toward the prior mean, particularly in years with sparse data.
With abundant data, the prior has less influence and the estimates converge toward the Maximum Likelihood values.

Rather than manually calculating intercept priors from published survival rates, `bboutools` provides functions to derive intercept priors from disturbance levels using national demographic-disturbance relationships.

## National Disturbance-Informed Priors

The user can specify priors informed by national demographic-disturbance relationships [@johnson_science_2020] instead of the default weakly informative priors.
These priors link anthropogenic and fire disturbance levels to expected survival and recruitment, based on the national boreal caribou model.

The functions `bb_priors_survival_national()` and `bb_priors_recruitment_national()` return intercept priors (`b0_mu` and `b0_sd`) based on the percent anthropogenic disturbance (`anthro`) and percent fire disturbance not overlapping with anthropogenic disturbance (`fire_excl_anthro`).
All other prior parameters retain their defaults from `bb_priors_survival()` and `bb_priors_recruitment()`.

```{r}
nat_priors <- bb_priors_survival_national(anthro = 50, fire_excl_anthro = 5)
nat_priors
```

For aggregate annual survival data, the `annual = TRUE` option returns priors on the annual survival scale.

```{r}
nat_priors_annual <- bb_priors_survival_national(
  anthro = 50,
  fire_excl_anthro = 5,
  annual = TRUE
)
nat_priors_annual
```

As an example, we compare survival estimates from the default priors and national disturbance-informed priors.

```{r}
set.seed(1)
data <- bboudata::bbousurv_c

fit_default <- bb_fit_survival(data, quiet = TRUE)

fit_national <- bb_fit_survival(
  data,
  priors = bb_priors_survival_national(anthro = 50, fire_excl_anthro = 5),
  quiet = TRUE
)
```

```{r, echo=FALSE,fig.width=6,fig.height=4}
pred_default <- bb_predict_survival(fit_default)
pred_national <- bb_predict_survival(fit_national)

df <- bind_rows(
  list(
    "Default Priors" = pred_default,
    "National (50% Anthro)" = pred_national
  ),
  .id = "model"
)

ggplot(data = df) +
  geom_pointrange(
    aes(
      x = factor(CaribouYear),
      y = estimate,
      ymin = lower,
      ymax = upper,
      color = model
    ),
    position = position_dodge(width = 0.5)
  ) +
  theme(legend.position = "bottom") +
  scale_y_continuous("Annual Survival", labels = scales::percent) +
  xlab("Caribou Year") +
  scale_color_discrete("")
```

Higher anthropogenic disturbance leads to lower expected survival.
The national priors shift the intercept accordingly, resulting in a more informative prior centered on the expected survival for the given disturbance level.

### Year Trend Prior and Data Rescaling

The `CaribouYear` variable is internally z-score standardized before model fitting.

$$Rescaled\_Year = \frac{CaribouYear - mean(CaribouYear)}{sd(CaribouYear)}$$

This rescaling affects the interpretation of the year trend prior `bYear`.
The prior `bYear ~ Normal(bYear\_mu, sd = bYear\_sd)` operates on the rescaled year scale, not the original calendar year scale.

The default `bYear_sd = 2` is intentionally diffuse on the rescaled scale and is generally appropriate for most datasets.

If a user wants to specify an informed trend prior per calendar year, the values must be converted to the rescaled scale by multiplying by the standard deviation of the years in the dataset.
For example, with years 2010--2015 (SD $\approx$ 1.87), a desired prior of 0.1 log-odds per calendar year with SD = 0.5 translates to `bYear_mu = 0.1 * 1.87` and `bYear_sd = 0.5 * 1.87` on the rescaled scale.

The rescaling improves numerical stability during model fitting and makes the default prior behave reasonably regardless of the absolute year values.
Both survival and recruitment models use identical rescaling.

## References
